Two subtle problems with over-representation analysis

### 2024-06-15 manuscript/manuscript.Rmd
mswork

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.

The lack of enrichment scores hampers interpretation.

Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis) [@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.

>>>>

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis) [@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.
